---
title             : "Exploring the Role of Deviance on Self-Concept Clarity Across the Lifespan"
shorttitle        : "Title"

author: 
  - name          : "Yan Wang"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "my@email.com"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"

affiliation:
  - id            : "1"
    institution   : "University of Chicago"

abstract: |
  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
  One sentence clearly stating the **general problem** being addressed by  this particular study.
  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
  One or two sentences to put the results into a more **general context**.
  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
  
  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : "keywords"
wordcount         : "X"

bibliography      : "SCC1.bib"

floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : papaja::apa6_pdf
figsintext        : yes
---

```{r setup, include = FALSE}
chooseCRANmirror(ind=65)
library("papaja")
#install.packages("correlation")
library(correlation)
#install.packages("Hmisc")
library(Hmisc)
#install.packages("lme4")
library(lme4)
#install.packages("haven")
library(haven)
#install.packages("magrittr")
library(magrittr)
#install.packages("readr")
library(readr)
#install.packages("dplyr")
library(dplyr)
#install.packages("knitr")
library(knitr)
#install.packages("ggplot2")
library(ggplot2)


library(tidyverse)
library(psych)
library(stats)
library(scales)
library(kableExtra)
r_refs("SCC1.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

Self-concept clarity means a clear and certain understanding of oneself and is usually associated with positive feelings about the self [@campbellSelfconceptClarityMeasurement1996; @campbellSelfesteemClaritySelfconcept1990] . In research on the self, self-concept clarity is one of the concepts associated closely with well-being [@campbellStructureSelfConceptIts2003]. Recently, a curvilinear relationship has been found on self-concept clarity over time [@lodi-smithGettingKnowMe2010]. Yet, the underlying mechanism for this relationship stays unclear. This study aims to test a potential explanation for the relationship between self-concept clarity and subjective well-being.

## Self-concept Clarity and Well-being

Researchers have found positive relationships between self-concept clarity and subjective well-being, emotional stability, resilience, and physical health; as well as negative relationships between self-concept clarity and negative affect, including anxiety and depression [@campbellSelfconceptClarityMeasurement1996; @campbellStructureSelfConceptIts2003; @lightSelfConceptClaritySelfRegulation2017; @ritchieSelfconceptClarityMediates2011;@lightInsOutsSelf2013;@stopaConstructingSelfRole2010]. Research also suggests that well-being factors can influence self-concept clarity: better well-being, especially physical health, relation stability, and role stability, is also associated with higher self-concept clarity, [@carterExaminingRelationshipSelfPerceptions2019; @slotterAllRoleTransitions2017]. It is notable that both well-being and self-concept clarity change with age.

## Self-Concept Clarity and Age
Despite the past assumption of a positive association between self-concept clarity and age, @lodi-smithGettingKnowMe2010 found a reverse U-shaped curvilinear relationship between the two variables across lifespan. But why the relationship between the two variables changes over time?
One potential explanation is social role. The entrance of a social role might add to self-concept therefore increasing self-concept clarity [@lightInsOutsSelf2013]. In contrast, the exit from a social role, especially when an individual is unprepared for the exit, may lead to a decrease in self-concept clarity [@lightInsOutsSelf2013]. When people are in emerging adulthood, even though they might exit their role as a student, they have other roles like employees, partnerships, parents, etc. to enter. For middle-aged people, their role entrance opportunities are more restricted than young adults and they might exit their roles as parents or children. However, generally speaking, their social roles are relatively stable and middle-aged adults are able to obtain and maintain social roles. Later adulthood is when social role exits start to rise. For example, retirement and independence of children can both lead to role exit. Without the cognitive attribution of pairing a role exit with entering a new role, the effect of role exit on self-concept clarity is associated with a drop in self-concept clarity [@lightInsOutsSelf2013]. 
In Light, in her -@lightUnpublishedData2024 unpublished data, they find a more specified interactive variable, the importance of self-concept clarity to subjective well-being, showing the same pattern as @lodi-smithGettingKnowMe2010's finding. Specifically, when the slope representing the effect of self-concept clarity on subjective well-being is on y-axis and age is on x-axis, the curve rises from emerging adult to middle age and drops in elderly adulthood, thus forming a reversed U-shape [@lightUnpublishedData2024].

## Well-being and Age
Another element to consider when asking why the relationship between self-concept clarity and well-being is curvilinear across the lifespan is the relationship between well-being and age. This topic is not only studied by psychologists but also by economists with parallel theories [@lopezulloaHowDoesSubjective2013]. Interestingly, not a cohesive conclusion has been made in both fields regarding what the shape of the slope is between well-being and age. The mixed result of linear, unchanging, U-shaped, and reversed U-shaped has been proposed and supported by different research in both fields [@lopezulloaHowDoesSubjective2013]. At this moment, U-shaped seems to be the most popular and accepted result in which well-being drops around middle age and rises back up during elderhood. According to the social roles’ influence above, it can be that although middle-aged adults have many stable social roles, not all of the roles are desirable, therefore having a lower overall well-being. However, most variations of the ongoing debate emerge in the data of more elderly adults [@blanchflowerHappinessUshapedEverywhere2021].
At the same time, social roles can be seen as part of the social norms, and social norms change for people at difference ages. Therefore, social roles may not be the full explanation for the found mechanism. In turn, I would like to propose social norms or social clock as an alternative explanation for this underlying mechanism for it is a more inclusive and holistic factor.

## Deviance and Well-being
Social clock consists of major life events happening within a normative age range [@helsonPersonalityPatternsAdherence1984]. Social clock theory states that being on-time has a positive association with well-being, whereas being off-time has a negative association to well-being [@helsonPersonalityPatternsAdherence1984; @neugartenTimeAgeLife1979]. It is commonly believed that being either early or late for major life-events can be stressful and leads to less social support. 
@rookTimingMajorLife1989 found that people who are late for a desirable life-event experience psychological distress but not people who are early. However, more current studies have found that individuals experience distress when experiencing life-events early [@abramsUnmetExpectationsWork2022; @pekel-uludagliYoungAdultsPerceptions2019]. For younger adults,@pekel-uludagliYoungAdultsPerceptions2019 found that people who perceived themselves as getting married and having children earlier than their peers experienced higher level of depression and lower level of need satisfaction. Their results support the social clock theory because people who perceived themselves as married on-time had higher life satisfaction and need satisfaction compared to both people who perceived themselves as late or early. Similarly, @abramsUnmetExpectationsWork2022 found that elderly adults who leave jobs unexpectedly have an increased risk of experiencing depressive symptoms. It is important to note that the perception of being off-time is often subjective in these study for the statistical average for life-events changes across generations.
Even though major life-events are about the same in a society, people from different age-groups might prioritize different concepts [@arnettConceptionsTransitionAdulthood2001]. For example, middle-aged people consider physical transition less important than emerging adults [@arnettConceptionsTransitionAdulthood2001]. Therefore, the perceived importance of norms might influence people’s perception on their self-concept, which in turn, may act as a mediator for the relationship between self-concept clarity and subjective well-being. In contrast, the perceived flexibility of norms at various life stages may moderate the strength of the relationship between self-concept clarity and subjective well-being. In other words, the extent to which people deviate from the social clock might be related to both their self-concept clarity and their well-being.

## Current Study
How can the role of deviance from the social clock explain the curvilinear relationship between self-concept clarity and subjective well-being? 
In this thesis, I intend to argue that 1) lower self-concept clarity will be associated with lower subjective well-being, 2) perceived deviance from the social clock will be associated with lower subjective well-being,3) perceived deviance from the social clock will mediate the association between self-concept clarity and subjective well-being, 4) perceived flexibility of the social clock will moderate the relationship between perceived deviance from the social clock and subjective well-being, 5) age will moderate the relationship between self-concept clarity & subjective well-being, and perceived deviance from the social clock & self-concept clarity.
-->

# Methods
```{r read-in data, include=FALSE}
SCC <- read_csv("Self-Concept Clarity and Deviance for R.csv")

SCCcor <- SCC[, c("SWLScr", "SCCScr", "MYA%", "MMA%", "MEA%", "Age")] %>%
  drop_na()

```

## Participants
Participants were recruited via Cloud Research. Cloud Research is an online platform for recruiting potential survey respondents. A total of `r nrow(SCCcor)` participants were recruited and included in the analysis (*M* = `r round(mean(SCCcor$Age), 2)`, *SD* = `r round(sd(SCCcor$Age), 2)`). All participants are American adults who can read English. 

## Material

### Self-Concept Clarity Scale
To measure participants’ level of self-concept clarity, participants completed the 12-item Self-Concept Clarity scale [@campbellSelfconceptClarityMeasurement1996]. Response options for this scale fall on a 7-point Likert scale ranging from 1 (Strongly Agree) to 7 (Strongly Disagree). An example item is “My beliefs about myself often conflict with one another.” Scale item scores were averaged and higher values reflect higher self-concept clarity.

### Satisfaction with Life Scale
To measure participants’ satisfaction with life, participants completed the 6-item  Satisfaction with Life scale [@dienerSatisfactionLifeScale1985]. Response options for this scale fall on a 7-point Likert scale ranging from 1 (Strongly Disagree) to 7 (Strongly Agree). An example item is “In most ways my life is close to my ideal.” Scale item scores were averaged and higher values reflect higher satisfaction of life.

### Self-Concept Clarity Norm Measures
To measure the perceived norm on self-concept clarity, participants completed the 6-item Self-Concept Clarity Norm Measures (developed by Dr. Light). The scale measures the descriptive and injunctive norms or belief about self-concept clarity. Response options for this scale fall on a 7-point Likert scale ranging from 1 (Strongly Disagree) to 7 (Strongly Agree). An example item for descriptive norms is “Most people my age have a clear sense of who and what they are.” An example item for injunctive norms is “Someone my age should have a clear and stable sense of self.” Scale item scores were averaged and higher values reflect higher perceived norm about self-concept clarity.

### Life-Events Markers
To measure if participants have reached certain life markers, I created a 41-item Life-Event Markers scale, partially adapted from the Marker of Adulthood items [@arnettConceptionsTransitionAdulthood2001] The items are life-events that are perceived as normative to complete throughout the life-span, including social, physical, and psychological markers. There are 16 items for younger adults (e.g. “Finished Education”); 13 items for middle-aged adults (e.g. “Interact with Parents as Equals”); and 12 items for elderly adults (e.g. “Been Able to Provide Advice to Younger Adults”). The order of the questions was randomized for the participants and, therefore involves no age-group bias. The instruction for the participants states “For each of the experiences listed, please let us know if this is one that you have personally experienced so far in your life.” Response options for this scale are forced-choice style questions with only “Yes” and “No”.
To measure the perceived deviance and flexibility of those normative life-markers, I took the 41-items in the Life Marker Scale and phrased them into sentences to ask the participants the percentage of people of their age’s life-markers completeness. The order of the questions was randomized for the participants and, therefore involves no age-group bias. The instruction for the participants states “Now we'd like you to reflect on the typical life experiences of people of your same age. Thinking about people of your same age, how many of them do you think have had each of the following life experiences? Please read the following items and answer the percentage of people your age who you think have experienced the life markers on a 100 percent scale.” One example item is “What percentage of people your age do you think have finished their education?” Response options for this scale are a slider scale ranging from 0% to 100% completeness in an interval of 1.
To analyze this scale, I first code “Yes” as 1 indicating the participant completed the life-marker, and “No” as - 1 indicating the participant has not completed the life-marker. Then, I multiply the 100% scale by their matching items. The resulting variable values ranged from -100 to +100 in intervals of 1, such that a score of -100 reflects participants who have not complete the life-marker and think that 100% of their peers have completed the life-marker and a score of +100 reflects participants who have complete the life-marker and think that 100% of the peers have completed the life-marker. Higher absolute value indicates less perceived flexibility and positive values smaller than 50 and negative values smaller than -50 indicate greater perceived deviance from the social clock.

### Future Time Perspective Scale
To measure participants’ perception of future time, participants completed the 10-item Future Time Perspective Scale (Carstensen & Lang, 1996). Response options for this scale fall on a 7-point Likert scale ranging from 1 (Very Untrue) to 7 (Very True). An example item is “Many opportunities await me in the future.” Scale item scores were averaged and higher values reflect greater future time perspective.

## Procedure
Participants first provided informed consent. Next, participants completed scales assessing self-concept clarity, satisfaction with life, self-concept clarity normative belief, future perspective, life-maker, and life-marker completeness of peers in randomized order. Finally, participants were prompted to provide their demographic information and were subsequently debriefed about the full purpose of the study, thanked for their participation, and financially compensated at the amount of $2 for completing the survey.

## Data analysis
We used `r cite_r("SCC1.bib")` for all our analyses.

# Results
The age distribution of the participants is shown in Figure \@ref(fig:agedis). With the minimal of 18 years old and the maximum of 86 years old.

```{r agedis, echo=FALSE,fig.cap="Age Distrubition", fig.width=4, fig.height=3}

#Histogram of age distributions
# SCC_Age <- SCC[, c("Age")] %>%
#   drop_na()
# ggplot(data = SCC_Age) +
#   geom_histogram(aes(x = Age), fill= "skyblue") +
#   scale_x_continuous(limits = c(18, 100), breaks = c(18, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90)) +
#   scale_y_continuous(limits = c(0, 100),breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100)) +
#   labs(x = "Age", 
#        y = "Count", 
#        title = "Age Distrubition") +
#   theme_minimal()+
#   theme(axis.title = element_text(family = "Times"),
#         panel.grid.minor = element_blank(),
#         axis.line = element_line(),
#         title = element_text(family = "Times"))

# Histogram of age distributions
SCC_Age <- SCC %>%
  select(Age) %>%
  na.omit()

ggplot(data = SCC_Age, aes(x = Age)) +
  geom_histogram(fill = "skyblue", binwidth = 3) +
  scale_x_continuous(breaks = seq(18, 86, by = 4)) +
  scale_y_continuous(limits = c(0, 95), breaks = seq(0, 95, by = 5)) +
  labs(x = "Age", 
       y = "Count", 
       title = "Age Distribution") +
  theme_minimal() +
  theme(axis.title = element_text(family = "Times"),
        panel.grid.minor = element_blank(),
        axis.line = element_line(),
        title = element_text(family = "Times"))

```


```{r cor, echo=FALSE, results='hide'}
#subdata of scc,swl,lms,age,age_sq
SCCcor <- SCC[, c("SWLScr", "SCCScr", "MYA%", "MMA%", "MEA%", "Age")] %>%
  drop_na()

#correlation matrix
matrix <- cor(SCCcor, method = "pearson")
####print(matrix)
cor_to_p(matrix, n = 659, method = "pearson")
```

```{r matrix, echo=FALSE, fig.width=6, fig.height=2}
#table
apa_table(matrix,
          caption = "Correlations Matrix for Self-Concept Clarity, Subjective Well-Being, and Life Marker Scale Across Age" ,
          note = "SWL = Satisfaction with Life Scale, SCC = Self-Concept Clairty Scale, YA = Life Marker Scale's Young Adult items score, MA = Life Marker Scale's Middle-age Adult items score, EA = Life Marker Scale's Elderly Adult items score")
#kable(matrix,
     # digits = 2,
      #col.names = c("SWL", "SCC", "YA", "MA", "EA", "Age" ),
      #align = c("c", "c", "c","c","c","c"),
      #caption = "Correlations Matrix for Self-Concept Clarity, Subjective Well-Being, and Life Marker Scale Across Age" ) %>%
 # kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  #kable_classic(html_font = "Times") %>%
  #row_spec(1, bold = T)
```
To test Hypothesis 1 and 2, I run a correlational analysis on Satisfaction with Life Scale, Self-Concept Clarity Scale, Life Marker Scale, and age, see in Table \@ref(tab:matrix). We found a significant positive association between self-concept clarity and satisfaction with life and all three age group's life marker degree of deviance positively associates with satisfaction with life. 

I further explored if age groups will influence the relationship between self-concept clarity and satisfaction with life (see Figure \@ref(fig:SCCtoSWL)).

```{r SCCtoSWL, echo=FALSE,fig.cap="Correlation between Self-Concept Clarity and Satisfaction with Life Across Age", fig.width=6, fig.height=3}
#Graph 1: SCC to SWL across three age group
SCC_SWL <- SCC[, c("SWLScr", "SCCScr","Age Group")] %>%
  drop_na()

ggplot(data = SCC_SWL, aes(x = SCCScr, y = SCCScr, color = `Age Group`)) +
  geom_smooth(method = lm) +
  geom_point() +
  scale_x_continuous(limits = , breaks = ) +
  scale_y_continuous(limits = ,breaks = ) +
  facet_wrap(~ SCC_SWL$`Age Group`, ncol = 3) +
  labs(x = "Self-Concept Clarity", 
       y = "Satisfaction with Life", 
       title = "Correlation between Self-Concept Clarity and Satisfaction with Life Across Age",
       color = "Age Groups") +
  theme_minimal()+
  theme(legend.position = "right",
        legend.title = element_text(family = "Times"),
        legend.text = element_text(family = "Times"),
        axis.title = element_text(family = "Times"),
        panel.grid.minor = element_blank(),
        axis.line = element_line())
```

# Discussion
The analysis provided evidence for Hypothesis 1 and 2 that self-concept clarity and degree of deviance from the social clock have a positive association with subjective well-being.

\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
